tstrait: a quantitative trait simulator for ancestral recombination graphs

### 2023-11-08 paper.tex
Part-way through intro

%\abstract{Abstracts must be able to stand alone and so cannot contain %citations to
%the paper's references, equations, etc. An abstract must consist of a single
%paragraph and be concise. Because of online formatting, abstracts must appear
%as plain as possible.}
\keywords{Ancestral Recombination Graph, ARG, GWAS, quantitative traits}

@@

Genome-wide association studies (GWAS)
identify genetic variants that are statistically
associated with a specific trait~\citep{uffelmann2021}.
GWAS has successfully
identified many loci that are associated with various human diseases and
traits~\citep{locke2015,ishigaki2022,mahajan2022,yengo2022,mathieson2023},
and GWAS results are actively being incorporated into clinical
practice~\citep{visscher2017}.
Despite these successes and the ongoing collection of population-scale
Biobank data~\citep{tanjo2021practical}, there are substantial open
questions around GWAS methology.
In particular, correctly accounting for population stratification
remains challenging~\citep{uffelmann2021}.

Ancestral recombination graphs (ARGs) describe the interwoven paths
of genetic inheritance arising from recombination~\citep{wong2023general}.
[TODO from here]
Recent studies have shown that genealogy-based association can detect more
ultra rare variants than conventional methods \citep{zhang2023}, using an
ARG-based approach in quantitative-trait locus mapping can better detect causal
locus, especially in the presence of allelic heterogeneity \citep{link2023},
and linkage disequilibirium can be efficiently modeled in GWAS by using a
tree-based approach \citep{salehi2023}. These recent advances make ARG to be a
suitable data format for conducting GWAS.

ARG is especially useful in population genetic simulations, and it would be
possible to simulate genetic data with ancestral information by using ARG.
Simulation is used in GWAS to examine how population structure and demographic
histories influence polygenic risk score prediction
\citep{martin2017,zaidi2020}, because simulation is the only way to obtain
genetic data where the underlying demographic information is known. A succinct
tree sequence (or tree sequence) is a data structure for ARGs, and it is one of
the most popular genomic data structures in population genetic simulations
\citep{baumdicker2022,haller2019}, (cite "A general and efficient, ..."
article). Since it can efficiently store genetic information with far less data
size than other methods, it is an appealing choice for simulating mutation and
ancestry information \citep{kelleher2019}.

@@

We believe that in the coming years, simulation in tree sequence encoding will
attract more users in need of conducting genetic simulations with complex
population structure. Multiple studies highlight the advantages of using the
ARG data structure in GWAS \citep{link2023,salehi2023,zhang2023}. ARG data
structure is becoming increasingly useful in genetic studies, as it is possible
to accurately infer biobank-scale genealogies from sequencing data
\citep{zhang2023}, and simulate large sample whole-genome sequences with
spatiotemporal metadata \citep{anderson2023}. These advances in ARG presents a
real opportunity for people to further understand the effects of complex
population structure on quantitative traits. The \texttt{tstrait} package is an
essential piece of infrastructure that lets people explore these possibilities.

>>>>

%\abstract{Abstracts must be able to stand alone and so cannot contain %citations to
%the paper's references, equations, etc. An abstract must consist of a single
%paragraph and be concise. Because of online formatting, abstracts must appear
%as plain as possible.}
\keywords{Ancestral Recombination Graph, ARG, GWAS, quantitative traits,
simulation}

@@

Genome-wide association studies (GWAS)
identify genetic variants that are statistically
associated with a specific trait~\citep{uffelmann2021}.
Many loci that are associated with various human diseases and
traits have been identified~\citep{locke2015,ishigaki2022,
mahajan2022,yengo2022,mathieson2023},
and GWAS results are actively being incorporated into clinical
practice~\citep{visscher2017}.
Despite these successes and the ongoing collection of population-scale
Biobank data~\citep{tanjo2021practical}, there are substantial open
questions around GWAS methology.
In particular, correctly accounting for population stratification
remain challenging~\citep{uffelmann2021}.

Ancestral recombination graphs (ARGs) encode the interwoven paths
of genetic inheritance arising from recombination~\citep{wong2023general}, and recent
breakthroughs have opened many new possibilities for their
practical application~\citep{lewanski2023era}.
Building on these breakthroughs, substantial progress has been made
on statistical genetics questions.
For example,  ARG-based methods
can detect more ultra rare variants than conventional association testing
methods~\citep{zhang2023};
can better detect causal loci in quantitative-trait locus mapping~\citep{link2023tree};
and can provide a sparse and efficient model of linkage disequilibirium
in GWAS and downstream applications~\citep{nowbandegani2023extremely}.

Simulation is a critical component of GWAS method development, and
generally consists of two steps: (1) generate a synthetic genetic
variation dataset; and (2) simulate some quantitative traits
based on this dataset.
(The two steps may sometimes be combined
in a single simulation, or we may use real data instead of simulations in
step (2), but the basic principles are still the same.)
The combined genetic variation and
simulated quantitative traits then represent the ``truth'' against
which association testing methods can be evaluated.
The first step, of generating realistic, large-scale genetic variation
datasets, is a well-studied problem with a mature and robust supporting
software ecosystem. ARG-based methods~\citep{kelleher2016efficient,
kelleher2018efficient,haller2018tree,baumdicker2022efficient,anderson2023}
have been instrumental in making simulations of the scale and realism
required to model modern GWAS cohorts~\cite[e.g.][]{martin2017,zaidi2020}.
These simulation methods are all based on the ``succinct tree
sequence'' ARG encoding~\citep{kelleher2019,ralph2020,wong2023general},
using the open-source \texttt{tskit} library.

% ARG is especially useful in population genetic simulations, and it would be
% possible to simulate genetic data with ancestral information by using ARG.
% Simulation is used in GWAS to examine how population structure and demographic
% histories influence polygenic risk score prediction
% \citep{martin2017,zaidi2020}, because simulation is the only way to obtain
% genetic data where the underlying demographic information is known. A succinct
% tree sequence (or tree sequence) is a data structure for ARGs, and it is one of
% the most popular genomic data structures in population genetic simulations
% \citep{baumdicker2022,haller2019}, (cite "A general and efficient, ..."
% article). Since it can efficiently store genetic information with far less data
% size than other methods, it is an appealing choice for simulating mutation and
% ancestry information \citep{kelleher2019}.

@@

We believe that in the coming years, simulation in tree sequence encoding will
attract more users in need of conducting genetic simulations with complex
population structure. Multiple studies highlight the advantages of using the
ARG data structure in GWAS \citep{link2023tree,nowbandegani2023extremely,zhang2023}.
ARG data
structure is becoming increasingly useful in genetic studies, as it is possible
to accurately infer biobank-scale genealogies from sequencing data
\citep{zhang2023}, and simulate large sample whole-genome sequences with
spatiotemporal metadata \citep{anderson2023}. These advances in ARG presents a
real opportunity for people to further understand the effects of complex
population structure on quantitative traits. The \texttt{tstrait} package is an
essential piece of infrastructure that lets people explore these possibilities.
